Finally Tagless, Partially Evaluated

### 2007-07-18 tagless/tagless-final.tex
Reinflating the paper for JFP: done up to Section 2.1

Our main idea is to encode
% de Bruijn or higher-order abstract syntax
HOAS
using cogen functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

\begin{quote}
\small
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \hfill\rm John Reynolds~\citep{reynolds-definitional}
\end{quote}

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
%aplas
\begin{comment}
No matter how we represent the object language in the
metalanguage, the representation can be created either by hand (for example, by
entering object terms at a metalanguage interpreter's prompt) or
by a parser\slash type\hyp checker reading from a text string.
We leave aside the solved problem of writing such a parser\slash type\hyp checker,
whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.
\end{comment}
%aplas-else
We leave aside the solved problem of writing a parser\slash type\hyp checker,
for embedding object language objects into the metalanguage
(whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}),
and just enter them by hand.
%aplas

@@

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.
It is straightforward to create an algebraic data type, say in OCaml~%
%aplas-specific
Fig.~\ref{fig:tag-problem}(a)%
, to
represent object terms such as those in Figure~\ref{fig:object}.
% fig:tag-problem part (a) was here
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let test1 = A (L (V VZ), B true)}.
%aplas-else
\begin{comment}
\begin{code}
let test1 = A (L (V VZ), B true)
\end{code}
\end{comment}

\begin{figure}
%
(a) \begin{code2}
type var = VZ | VS of var
type exp = V of var | B of bool | L of exp | A of exp * exp
\end{code2}

(b) \begin{code2}
let rec lookup (x::env) = function | VZ   -> x
                                   | VS v -> lookup env v
let rec eval0 env = function
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{code2}

(c) \begin{code2}
type u = UB of bool | UA of (u -> u)
\end{code2}

(d) \begin{code2}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{code2}
\caption{OCaml code illustrating the tag problem.}
\label{fig:tag-problem}
\end{figure}
% 
Following \cite{WalidICFP02},
we try to implement an interpreter function |eval0|%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(b)%
. It takes
an object term such as |test1| above and gives us its value.
The first argument to |eval0| is the environment, initially empty,
which is the list of values bound to free variables in the
interpreted code.
% fig:tag-problem part (b) was here
If our OCaml-like metalanguage were untyped, the code above would be 
acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \citep[\S1.1--2]{WalidICFP02}.

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(c)%
.
% fig:tag-problem part (c) was here
We can then write a typed interpreter%
%aplas-specific
, Fig.~\ref{fig:tag-problem}(d)
% fig:tag-problem part (d) was here
whose inferred type is |u list -> exp -> u|. Now we can evaluate
%aplas-specific
\texttt{eval [] test1} obtaining |UB true|.
%aplas-else
\begin{comment}
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}
\end{comment}
% fig:tag-problem part (f) was here
The unfortunate tag |UB| in the result reflects that |eval| is a partial
function.  
%aplas-specific
First, the pattern match |with UA f| in the line |A (e1,e2)| is not 
exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term |A (L (V (VS VZ)), B true)|.
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.
%aplas-else
\begin{comment}
First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term
\begin{code}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.
\end{comment}

@@

%aplas-else
\begin{comment}
\subsection{Solutions using fancier types}
\end{comment}

It is commonly thought that to interpret a typed object language in
a typed metalanguage while preserving types is difficult and requires
GADTs or dependent types \citep{taha-tag}.  In fact, this problem
motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
%aplas-else
\begin{comment}
For a metalanguage's type system to allow the well-typed object term
|test1| but disallow the ill-typed object term |test2|, fancier types
such as GADTs or dependent types seem necessary.  
\end{comment}
Yet other type systems
have been proposed to distinguish closed terms like |test1| from open
terms like |test3|
%aplas \cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual},
\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01},
so that |lookup| never receives an empty environment.  We discuss these
proposals further in \S\ref{related}%
%aplas-specific
.
%aplas-else
\begin{comment}
; here we just note that many
advanced type systems have been devised to ensure statically that an
object term is well-typed and closed.
\end{comment}

@@

We represent object programs using ordinary functions rather than
data constructors.  The interpreter below provides these functions
\begin{code4}
let varZ env    = fst env       let b (bv:bool) env = bv
let varS vp env = vp (snd env)  let app e1 e2 env   = (e1 env) (e2 env)
let lam e env   = fun x -> e (x,env)
\end{code4}
which comprise the entire interpreter.
We now represent our sample term $(\fun{x}x)\True$ as
%aplas-specific
\texttt{let testf1 = app (lam varZ) (b true)}
%aplas-else
\begin{comment}
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
\end{comment}
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment%
%aplas-specific
, |testf1 ()|, obtaining |true|.
%aplas-else
\begin{comment}
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
\end{comment}
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
%aplas-specific
open term such as \texttt{app (lam (varS varZ)) (b true)}
gives a type- rather than a run-time error.
%aplas-else
\begin{comment}
\begin{code}
open term such as |testf3| below gives a type error rather than
a run-time error
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
\end{comment}
The type error correctly complains
that the initial environment should be a tuple rather than |()|.
In other words, the term is open.

In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
%aplas
Hindley-Milner type system
%aplas \cite{hindley-principal,milner-theory}
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

Our solution is emphatically distinct from Church-encoding of the
universal type. Church-encoding of the type |u|, \S\ref{tagproblem},
requires two continuations; the interpreter (see |app| above) would
have to provide both to the encoding of |e1|. The continuation
corresponding to the |UB| branch of |u| must either raise an error or
loop. For a well-typed object term that error continuation is never
invoked, yet it must be supplied. Our interpreter has no error
continuation at all.

The evaluator above is wired directly into the
functions |b|, |lam|, |app|, and so on.  %In the rest of this paper, 
We explain how to abstract the interpreter so as
to process the \emph{same} term in many other
ways: compilation, partial evaluation, CPS
conversion, and so forth.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage
%aplas
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}%
.
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\cite{metaocaml}) and standard Haskell. The complete code is available
\url{\oleg{put URL!}.}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

We take the problem of tagless (staged) typed-preserving
interpretation exactly as it was posed in \cite{WalidICFP02} and
\cite{xi-guarded}.  We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles%
% aplas
%   (\S\ref{variations}).
.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation%
% aplas: commented
%  and CPS transformation (\S\ref{PE}).
.
\item We point a clear way to extend the object language with more features
    such as state%
%aplas (\S\ref{state})%
%aplas
\footnote{again, see code}
.
\item We describe an approach to self\hyp interpretation compatible with the
  above%
%aplas: (\S\ref{selfinterp})
.  Self\hyp interpretation turned out
  harder than expected%
%aplas
\footnotemark[\value{footnote}]
.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

We reiterate that we specifically do not propose any new language
feature or technique. We use the existing, known approaches --
Hindley-Milner type system with inference-preserving module system or
constructor classes extensions, as realized in ML or Haskell98 -- to
solve the problem that was stated in published record as lacking
solution in ML or Haskell98 and most likely unimplementable in these
languages without extensions. The simplicity of our solution and our
use of only mainstream functional language features makes typed
embedded DSLs more practical to build.
\begin{comment}
We may claim some contribution about type-preserving CPS. Check
related work section in \cite{Guillemette-Monier-PLPV}, especially
check the work of Shao on type-preserving CPS in Flint. The PLPV paper
in related work shows other tasks, including closure conversion, which
we may tackle in our approach. We may be able to write a
type-preserving, assured compiler, whose properties and assured by HM.
\end{comment}

@@

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)%
%aplas-specific
, Fig.~\ref{fig:symantics-haskell}(a)%
. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\begin{figure}
%
(a) \begin{code2}
class Symantics repr where
  int  :: Int  -> repr Int       bool :: Bool -> repr Bool
  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

  add :: repr Int -> repr Int -> repr Int
  mul :: repr Int -> repr Int -> repr Int
  leq :: repr Int -> repr Int -> repr Bool
  if_ :: repr Bool -> repr a -> repr a -> repr a
\end{code2}

(b) \begin{code2}
testpowfix () = lam (\x -> fix (\self -> lam (\n ->
                 if_ (leq n (int 0)) (int 1)
                     (mul x (app self (add n (int (-1))))))))
\end{code2}

(c) \begin{code2}
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code2}
\caption{Symantics in Haskell.}
\label{fig:symantics-haskell}
\end{figure}
% fig:symantics-haskell part (a) was here
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
whose inferred type is \texttt{Symantics repr => repr Bool}.
For another example, the classical $\mathit{power}$ function is
%aplas-specific
in Fig.~\ref{fig:symantics-haskell}(b)
% fig:symantics-haskell part (b) was here
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
in Fig.~\ref{fig:symantics-haskell}(c).
% fig:symantics-haskell part (c) was here
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag.
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete implementations.

@@

  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> ('c, 'da -> 'db) repr
  val app : ('c, 'da -> 'db) repr -> ('c, 'da) repr -> ('c, 'db) repr
  val fix : (('c, 'da -> 'db) repr -> ('c, 'da -> 'db) repr) -> ('c, 'da -> 'db) repr

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr-> (unit-> ('c, 'da) repr)-> (unit-> ('c, 'da) repr)-> ('c, 'da) repr
end

>>>>

Our main idea is to encode
\ifshort HOAS \else de Bruijn or higher-order abstract syntax \fi
using cogen functions rather than data constructors.
In other words, we represent object terms not in an initial algebra
but using the coalgebraic structure of the $\lambda$-calculus.
Our representation also simulates inductive maps from types to
types, which are required for typed partial evaluation and CPS transformations.

@@

\begin{quote}
\small
    It should also be possible to define languages with a highly
    refined syntactic type structure. Ideally, such a treatment should
    be metacircular, in the sense that the type structure used in the
    defined language should be adequate for the defining language.
    \rm \ifshort \hfill John Reynolds~\fi\citep{reynolds-definitional}
\end{quote}

@@

\Citet{WalidICFP02} and \citet{xi-guarded} motivated interpreting
a typed object language in a typed metalanguage as an interesting
problem.  The known solutions to this problem store object terms and
values in the metalanguage in a universal type, a generalized algebraic
data type (GADT), or a dependent type.  In the remainder of this section,
we discuss these solutions, identify their drawbacks, then summarize our
proposal and contributions.  
\ifshort
We leave aside the solved problem of writing a parser\slash type\hyp checker,
for embedding object language objects into the metalanguage
(whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}),
and just enter them by hand.
\else
No matter how we represent the object language in the
metalanguage, the representation can be created either by hand (for example, by
entering object terms at a metalanguage interpreter's prompt) or
by a parser\slash type\hyp checker reading from a text string.
We leave aside the solved problem of writing such a parser\slash type\hyp checker,
whether using dependent types \cite{WalidICFP02} or not \cite{baars-typing}.
\fi

@@

\begin{SaveVerbatim}{2a}
type var = VZ | VS of var
type exp = V of var | B of bool | L of exp | A of exp * exp
\end{SaveVerbatim}
\begin{SaveVerbatim}[commandchars=\@\{\}]{2b}
let rec lookup (x::env) = function VZ -> x | VS v -> lookup env v
let rec eval0 env = function
| V v       -> lookup env v
| B b       -> b 
| L e       -> fun x -> eval0 (x::env) e
| A (e1,e2) -> (eval0 env e1) (eval0 env e2) 
\end{SaveVerbatim}
\begin{SaveVerbatim}{2c}
type u = UB of bool | UA of (u -> u)
\end{SaveVerbatim}
\begin{SaveVerbatim}{2d}
let rec eval env = function
| V v       -> lookup env v
| B b       -> UB b
| L e       -> UA (fun x -> eval (x::env) e)
| A (e1,e2) -> match eval env e1 with UA f -> f (eval env e2)
\end{SaveVerbatim}
\begin{SaveVerbatim}{test1}
let test1 = A (L (V VZ), B true)
\end{SaveVerbatim}

% see tagless\_interp1.ml, module Tagfull for the complete code.
% If you change the code in here, please adjust the .ml file
% accordingly. Let the paper and the accompanying code be in sync.
It is straightforward to create an algebraic data type, say in OCaml\ifshort\
Fig.~\ref{fig:tag-problem}(a)\fi, to
represent object terms such as those in Figure~\ref{fig:object}.
For brevity, we elide treating integers, conditionals, and fixpoint in
this section.
\ifshort\else\UseVerbatim{2a}\fi
We represent each variable using a unary de Bruijn index.
For example, we represent the object term $(\fun{x}x)\True$ as
\ifshort \BUseVerbatim{test1}.\else \UseVerbatim{test1}\fi

\ifshort
\begin{figure}
%
(a) \BUseVerbatim{2a}

(b) \BUseVerbatim{2b}

(c) \BUseVerbatim{2c}

(d) \BUseVerbatim{2d}
\caption{OCaml code illustrating the tag problem.}
\label{fig:tag-problem}
\end{figure}
\fi

Following \cite{WalidICFP02},
we try to implement an interpreter function |eval0|\ifshort,
Fig.~\ref{fig:tag-problem}(b)\fi. It takes
an object term such as |test1| above and gives us its value.
The first argument to |eval0| is the environment, initially empty,
which is the list of values bound to free variables in the
interpreted code.
\ifshort\else\UseVerbatim{2b}\fi
If our OCaml-like metalanguage were untyped, the code above would be 
acceptable.
The |L e| line exhibits interpretive overhead:
|eval0| traverses the function body~|e| every time (the result of
evaluating) |L e| is applied. Staging can be used to remove this
interpretive overhead \citep[\S1.1--2]{WalidICFP02}.

@@

The usual solution is to introduce a universal type \citep[\S1.3]
{WalidICFP02} containing both booleans and functions\ifshort,
Fig.~\ref{fig:tag-problem}(c)\fi.
\ifshort\else\UseVerbatim{2c}\fi
We can then write a typed interpreter\ifshort,
Fig.~\ref{fig:tag-problem}(d), \else\UseVerbatim{2d}\fi
whose inferred type is |u list -> exp -> u|. Now we can evaluate
\ifshort
\texttt{eval [] test1} obtaining |UB true|.
\else
\begin{code}
let test1r = eval [] test1
val test1r : u = UB true 
\end{code}
\fi
The unfortunate tag |UB| in the result reflects that |eval| is a partial
function.  
First, the pattern match |with UA f| in the line
|A (e1,e2)| is not exhaustive, so |eval| can fail if we apply a boolean,
as in the ill-typed term |A (B true, B false)|.
\ifshort\else
\begin{code}
let test2 = A (B true, B false)
let test2r = eval [] test2
Exception: Match_failure in eval
\end{code}
\fi
Second, the |lookup|
function assumes a nonempty environment, so |eval| can fail if we
evaluate an open term
\ifshort
|A (L (V (VS VZ)), B true)|.
\else
\begin{code}
let test3 = A (L (V (VS VZ)), B true)
let test3r = eval [] test3
Exception: Match_failure in lookup
\end{code}
\fi
After all, the type |exp| represents object
terms both well-typed and ill-typed, both open and closed.

@@

\ifshort\else\subsection{Solutions using fancier types}\fi

It is commonly thought that to interpret a typed object language in
a typed metalanguage while preserving types is difficult and requires
GADTs or dependent types \citep{taha-tag}.  In fact, this problem
motivated much work on GADTs \cite{xi-guarded,peyton-jones-simple} and
on dependent types \cite{WalidICFP02,fogarty-concoqtion}.
\ifshort\else
For a metalanguage's type system to allow the well-typed object term
|test1| but disallow the ill-typed object term |test2|, fancier types
such as GADTs or dependent types seem necessary.  
\fi
Yet other type systems
have been proposed to distinguish closed terms like |test1| from open
terms 
\ifshort\cite{WalidPOPL03,NanevskiJFP05,DaviesJACM01}\else
like |test3|
\cite{WalidPOPL03,NanevskiICFP02,NanevskiJFP05,DaviesJACM01,nanevski-contextual}\fi,
so that |lookup| never receives an empty environment.  We discuss these
proposals further in \S\ref{related}\ifshort\else;
here we just note that many
advanced type systems have been devised to ensure statically that an
object term is well-typed and closed\fi.

@@

We represent object programs using ordinary functions rather than
data constructors.  These functions comprise the entire interpreter, shown
below.
\ifshort
\begin{code3}[xleftmargin=0pt]
let varZ env    = fst env       let b (bv:bool) env = bv
let varS vp env = vp (snd env)  let app e1 e2 env   = (e1 env) (e2 env)
let lam e env   = fun x -> e (x,env)
\end{code3}
\else
\begin{code}
let varZ env        = fst env
let varS vp env     = vp (snd env)
let b (bv:bool) env = bv
let lam e env       = fun x -> e (x,env)
let app e1 e2 env   = (e1 env) (e2 env)
\end{code}
\fi
We now represent our sample term $(\fun{x}x)\True$ as
\ifshort
\texttt{let testf1 = app (lam varZ) (b true)}.
\else
\begin{code}
let testf1 = app (lam varZ) (b true)
\end{code}
\fi
This representation is almost the same as in \S\ref{tagproblem}, only
written with lowercase identifiers. To evaluate an object term is to
apply its representation to the empty environment\ifshort
, |testf1 ()|, obtaining |true|\fi.
\ifshort\else
\begin{code}
let testf1r = testf1 ()
val testf1r : bool = true
\end{code}
\fi
The result has no tags: the interpreter patently uses no tags and no
pattern matching. The term |b true| evaluates to a boolean and the term
|lam varZ| evaluates to a function, both untagged. The |app| function
applies |lam varZ| without pattern matching. What is more, evaluating an
open term such as
\ifshort
\texttt{app (lam (varS varZ)) (b true)}
\else
|testf3| below
\fi
gives a type error rather than a run-time error.
\ifshort\else
\begin{code}
let testf3 = app (lam (varS varZ)) (b true)
let testf3r = testf3 ()
\end{code}
\fi
The type error correctly complains
that the initial environment should be a tuple rather than~|()|.
In other words, the term is open.

In sum, by Church\hyp encoding terms using ordinary functions, we
achieve a tagless evaluator for a typed object language in a
metalanguage with a simple 
\ifshort
Hindley-Milner \fi
type system\ifshort\else\ \cite{hindley-principal,milner-theory}\fi
.  In this \emph{final} rather
than \emph{initial} approach, both kinds of run-time errors in
\S\ref{tagproblem} (applying a nonfunction and evaluating an open
term) are reported at compile time. Because the new interpreter
uses no universal type or pattern matching, it never results in a
run-time error (and is in fact total).  Because this safety is obvious
not just to us but also to the metalanguage implementation, we avoid
the serious performance penalty \cite{WalidICFP02} of error checking.
\Citet{Gluck-jones-optimality} explains deeper technical reasons that
inevitably lead to these performance penalties.

Our solution is \emph{not} Church-encoding the
universal type. The Church encoding of the type~|u| in \S\ref{tagproblem}
requires two continuations; the function |app| in the interpreter above would
have to provide both to the encoding of~|e1|. The continuation
corresponding to the |UB| case of~|u| must either raise an error or
loop. For a well-typed object term, that error continuation is never
invoked, yet it must be supplied. In contrast, our interpreter has no error
continuation at all.

The evaluator above is wired directly into the
functions |b|, |lam|, |app|, and so on.  \ifshort We \else In the rest of this paper, we \fi
explain how to abstract the interpreter so as
to process the \emph{same} term in many other
ways: compilation, partial evaluation, CPS
conversion, and so forth.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN) CPS, so to isolate the evaluation
        order of the object language from that of the metalanguage.\ifshort
\footnote{Due to serious lack of space, 
we refer the reader to the accompanying code for this.}\fi
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
MetaOCaml) and standard Haskell. The complete code is available at
\url{http://okmij.org/ftp/packages/tagless-final.tar.gz}
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

We attack the problem of tagless (staged) typed-preserving
interpretation exactly as it was posed
by \citet{WalidICFP02} and \citet{xi-guarded}.
We use their running examples and achieve the
result they call desirable.  Our contributions are as follows.

\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned out
  harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\cite{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

To reiterate, we do \emph{not} propose any new language
feature or technique. We use features already present in
mainstream functional languages---Hindley-Milner
type system with either an inference\hyp preserving module system or
constructor classes, as realized in ML and Haskell 98---to
solve the problem that was stated in the published record as
unsolved and likely unsolvable in ML or Haskell 98
without extensions. The simplicity of our solution and our
use of only mainstream features makes it more practical to build typed,
embedded domain\hyp specific languages.
\begin{comment}
We may claim some contribution about type-preserving CPS. Check
related work section in \cite{Guillemette-Monier-PLPV}, especially
check the work of Shao on type-preserving CPS in Flint. The PLPV paper
in related work shows other tasks, including closure conversion, which
we may tackle in our approach. We may be able to write a
type-preserving, assured compiler, whose properties and assured by HM.
\end{comment}

@@

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}

\ifshort
\begin{SaveVerbatim}{3a}
class Symantics repr where
  int :: Int  -> repr Int;       bool :: Bool -> repr Bool
  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

  add :: repr Int -> repr Int -> repr Int
  mul :: repr Int -> repr Int -> repr Int
  leq :: repr Int -> repr Int -> repr Bool
  if_ :: repr Bool -> repr a -> repr a -> repr a
\end{SaveVerbatim}
\else
\begin{SaveVerbatim}{3a}
class Symantics repr where
  int  :: Int  -> repr Int
  bool :: Bool -> repr Bool

  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

  add :: repr Int -> repr Int -> repr Int
  mul :: repr Int -> repr Int -> repr Int
  leq :: repr Int -> repr Int -> repr Bool
  if_ :: repr Bool -> repr a -> repr a -> repr a
\end{SaveVerbatim}
\fi
\begin{SaveVerbatim}{3b}
testpowfix () = lam (\x -> fix (\self -> lam (\n ->
                 if_ (leq n (int 0)) (int 1)
                     (mul x (app self (add n (int (-1))))))))
\end{SaveVerbatim}
\begin{SaveVerbatim}{3c}
testpowfix7 () = lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{SaveVerbatim}

We embed our language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics| (with a parameter |repr| of kind |* -> *|)\ifshort,
Fig.~\ref{fig:symantics-haskell}(a)\fi. The class is so named
because its interface gives the syntax of the object language and its
instances give the semantics.
\ifshort\else\UseVerbatim{3a}\fi
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as \texttt{app (lam (\textbackslash x -> x)) (bool True)},
whose inferred type is \texttt{Symantics repr => repr Bool}.
For another example, the classical $\mathit{power}$ function is
\ifshort in Fig.~\ref{fig:symantics-haskell}(b)
\else\UseVerbatim{3b}\fi
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
\ifshort in Fig.~\ref{fig:symantics-haskell}(c).
\else\UseVerbatim{3c}\fi
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomorphism restriction
with a compiler flag.
The methods |add|, |mul|, and |leq| are quite similar, and so are
|int| and |bool|. Therefore, we will frequently show implementations of
only one method of each group and elide the rest, to save space. The
accompanying code has the complete implementations.

\ifshort
\begin{figure}
(a) \BUseVerbatim{3a}

(b) \BUseVerbatim{3b}

(c) \BUseVerbatim{3c}
\caption{Symantics in Haskell}
\label{fig:symantics-haskell}
\end{figure}
\fi

@@

  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> ('c, 'da -> 'db) repr
  val app : ('c, 'da -> 'db) repr -> ('c, 'da) repr -> ('c, 'db) repr
  val fix : ('x -> 'x) -> (('c, 'da -> 'db) repr as 'x)
